---
title: "SpeechProject"
author: "Aditya Maheshwari"
date: '2018-08-10'
output: html_document
---

# Introduction

The goal is to authenticate a device or open a lock using voice recognition. For now we are fixing a phrase to be used as the password ("open the door"). We will use 50 different voices just to demonstrate a proof of concept. The function created will check two things:

+ 1: The phrase spoken matches the actual password phrase.

For that we will use the google speech to text wrapped in a package called googleLanguageR. This requires a mono wav file which for now we are converting with audacity, so it will take an extra step. 

```{r speechToText}
#install.packages("googleLanguageR")
library(googleLanguageR)

filepathTrack <- "~/LocalDocs/MLNotebooks/h.wav"
filepathAuth <- "~/LocalDocs/MLNotebooks/speechToText-b4ef08c4fe96.json"
gl_auth(filepathAuth)
gl_speech(filepathTrack, sampleRateHertz=48000)
```

+ 2: The frequency and tone of the voice matches that of the password

For that we will use the tuneR function to translate the voice into a vector of frequencies. From there we will use a combination of machine learning and statistical measures to determine how far the voice is from the authenticated voice. 

```{r frequencies}
#install.packages("seewave")
library(seewave)

filepathOutput <- "~/LocalDocs/MLNotebooks/txtTrack.txt"
myTrack<-readWave(filepathTrack)

export(myTrack, f=48000, filename=filepathOutput)

newDf <- read.delim(filepathOutput)
colnames(newDf) <- c("frequency")

getmode <- function(v) {
   uniqv <- unique(v)
   mode <- uniqv[which.max(tabulate(match(v, uniqv)))]
   print(length(v[v == mode]))
   mode
}

getmodes <- function(v, aRange = 1:5) {
    checkV <- v
    modes <- c()
    for (i in aRange) {
        modes <- c(modes, getmode(checkV))
        checkV <- checkV[checkV != modes[i]]
    }
    modes
}

# newVec <- apply(vec, FUN = mean(vec[]))
findMeans <- function(v) {
  apply(v[
}

```
